---
title: "MULTIVARIATE ANALYSIS -- H1B VISA DATA -- Exploratory Data Analysis (EDA)"
author: "Ashita Shetty"
output:
  html_document: default
  pdf_document: default
---


### Loading the data 

```{r}
library(readr)
h1b <- read_csv("C:\\Users\\Dell\\Desktop\\RBS\\Semester II\\Multivariate Analysis\\Assignment\\data\\H1_B-2014.csv")
str(h1b)

```

### Loading packages

```{r}
library(lattice)
library(ggplot2)
library(ggridges)
library(ggvis)
library(ggthemes)
library(cowplot)
library(gapminder)
library(gganimate)
library(dplyr)
library(tidyverse)
library(grid)
library(gridExtra)
library(RColorBrewer)

options(scipen=999)
```

### EDA



##### **1. What is the distribution of visa applicant status?**

```{r}
ggplot(h1b, aes(x = STATUS, fill = STATUS)) +
  geom_bar() +
  labs(title = "Distribution of Visa Status",
       x = "Status",
       y = "Count") + 
  scale_fill_manual(values = c("CERTIFIED" = "steelblue", "REJECTED" = "red"))+
  theme_fivethirtyeight()

```

##### **Inference:**
It can be observed that *majority of the applicants have been approved of their visa application status*.



##### **2. What is the distribution of Visa Classes based on their approval status?**

```{r}
ggplot(h1b, aes(VISA_CLASS)) + facet_grid(.~STATUS) + 
  geom_bar(fill="black", position="stack", width = 0.8) + coord_flip()+ theme_fivethirtyeight()
```

##### **Inference:**
* The analysis suggests that the majority of applicants opted for H1B, with E-3 Australian being the next popular choice.
* Notably, **H1B** stands out as having the highest number of certifications, indicating a predominantly positive status.



##### **3. Try to understand the Wage rate for case applicants, and dig further into the segregation based on Visa applied for**

```{r}
summary(h1b$LCA_CASE_WAGE_RATE_FROM)
```


```{r}
ggplot(h1b, aes(x = VISA_CLASS, y = LCA_CASE_WAGE_RATE_FROM)) + 
  geom_boxplot() + coord_flip()+
  theme_fivethirtyeight()
```

##### **Inference:** 
* The evidence demonstrates that the median amounts for most visa types are approximately within a similar range.
* However, an outlier is noticeable in the H-1B visa type, where the maximum value exceeds 800,000.



##### **4. What is the trend in the decision period and is there a pattern to be observed that exposes when most decisions are usually out?**

```{r}
library(lubridate)  
# Convert DECISION_DATE to a datetime object
h1b$DECISION_DATE <- dmy_hm(h1b$DECISION_DATE)

# Extract year and month from the datetime
h1b <- h1b %>% mutate(YearMonth = format(DECISION_DATE, "%Y-%m"))
```

```{r}
ggplot(h1b, aes(x = YearMonth, group = 1)) + 
  geom_line(stat = "count") + 
  labs(title = "Time Trends of Visa Application Decisions", x = "Year-Month",y = "Number of Visa Applications")+
  theme_fivethirtyeight()
```

##### **Inference:**
* It is noticeable that the majority of results were announced in March for the year 2014.
* Following March, there was a significant decrease, averaging around 1,500, which was also observed before March.



##### **5. What are the 5 Most Common Job titles for H1B Visa Holders?**

```{r}
h1b %>%
  count(LCA_CASE_JOB_TITLE) %>%
  top_n(5, n) %>%
  ggplot(aes(x = reorder(LCA_CASE_JOB_TITLE, n), y = n, fill = LCA_CASE_JOB_TITLE)) +
  geom_bar(stat = "identity") +
  labs(title = "5 Most Common Job Titles for H-1B Visa Holders",
       x = "Job Title",
       y = "Frequency") +
  theme_fivethirtyeight() + coord_flip()+ theme(legend.position = "none", plot.title.position = "plot")

options(repr.plot.width = 50)

```

##### **Inference:**
* The visual features the top 5 job roles of H1B Visa Holders.
* **Programmer Analyst is a clear winner in this!**



##### **6. Comparison between Wage Rates & Prevailing Wage**

```{r}
ggplot(h1b, aes(x = LCA_CASE_WAGE_RATE_FROM, y = PW_1)) +
  facet_wrap(.~STATUS)+
  geom_point() +
  labs(title = "Scatter Plot of Wage Rates vs. Prevailing Wage",
       x = "Wage Rate (From)",
       y = "Prevailing Wage") +
  theme_fivethirtyeight()

```

##### **Inference:**
* The examination of the scatter plot indicates a robust positive correlation due to the concentrated grouping of most points along a rising trend.
* Some outliers are observed, particularly in Rejected cases.
* A linear relationship is evident between the two variables.



#### **7. What is the correlation between Numerical columns in the data?**

```{r}
h1b$WAGE_RATE_FROM <- as.numeric(h1b$LCA_CASE_WAGE_RATE_FROM)
h1b$EMPLOYER_POSTAL_CODE <- as.numeric(h1b$LCA_CASE_EMPLOYER_POSTAL_CODE)
h1b$EMP_DIFF <- as.numeric(h1b$EMP_DIFF)
h1b$APP_DIFF <- as.numeric(h1b$APP_DIFF)
h1b$TOTAL_WORKERS <- as.numeric(h1b$TOTAL_WORKERS)
h1b$YR_SOURCE <- as.numeric(h1b$YR_SOURCE_PUB_1)
h1b$NAICS_CODE <- as.numeric(h1b$LCA_CASE_NAICS_CODE)
```


```{r}
library(corrplot)

cor_matrix <- cor(h1b[, c("WAGE_RATE_FROM", "EMPLOYER_POSTAL_CODE", "APP_DIFF", "EMP_DIFF", "TOTAL_WORKERS", "YR_SOURCE", "NAICS_CODE")])

corrplot(cor_matrix, type = "upper", method = "color")
```

##### **Inference:**
* The majority of these columns exhibit no correlations, with a few displaying subtle positive associations.
* Examples of columns with positive correlations include: Wage rate and employer postal code; Employer Postal Code and the applicant's tenure in the company; Likewise, the applicant's tenure in the company and the total number of workers in that specific company.
* There is a negative correlation between the year of introduction of the prevailing wage source (YR_SOURCE) and the time difference between the applicant's case date and decision date (APP_DIFF).

